---
title: "Using shape files with R & Python"
author: "Alexander Rodionov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using shape files with R & Python}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```

```{r setup}
packages <- c("devtools", "testthat", "dplyr", "ggplot2", "sf", "magrittr", "RColorBrewer", "leaflet", "tmap", "usethis", "reticulate")
utils::install.packages(base::setdiff(packages, base::rownames(utils::installed.packages())))   # R ver. 3.6.2

library(geokz)
```

`geokz`-package provides access to multiple dataset of different types and for different use. In this vignette we introduce the different datas and explain their use cases. Vignette *Using shape files with R & Python* provides multiple real-world examples using shapefiles with **R** and **Python**.

## Package installation {#package_installation}

`geokz` can be installed from GitHUB using.

```{r eval=FALSE, include=FALSE}
# Install development version from GitHub
devtools::install_github("arodionoff/geokz")

# Install development version from GitHub with vignette
devtools::install_github("arodionoff/geokz", build_vignettes = TRUE)
```

## Load shape file with Administrative units level 1

Kazakhstan is divided into 20 regions (Kazakh: облыстар/oblystar; singular: облыс/oblys; Russian: области/oblasti; singular: область/oblast' ).

```{r kaz_adm1_sf}
library(geokz)
library(dplyr)
library(sf)

IQR <- data(package = "geokz")
dplyr::as_tibble(IQR$results) %>% 
  dplyr::select(Item, Title) %>% 
    dplyr::filter(row_number() != n()) %>% # drop on `natural_zones` w/o shapefile
      dplyr::transmute(File = list.files(path = system.file("shape", package = "geokz"), pattern = "*.shp"),
                       Item, Title)

shpfile <- 
  system.file("shape", "kaz_admbnda_adm1_2024.shp", package = "geokz")

# agr - attribute-geometry-relationship, specifies for each non-geometry attribute column how it relates to the geometry
agr = c(KATO = "identity", ADM0_EN = "constant", ADM0_KK = "constant", ADM0_RU = "constant", ADM0_PCODE = "constant",
        ADM1_EN = "identity", ADM1_KK = "identity", ADM1_RU = "identity", ADM1_PCODE = "identity",
        ISO_3166_2 = "identity")

kaz_ADM1_sf <- 
  sf::st_read(dsn = shpfile, agr = agr)

print(sf::st_crs(kaz_ADM1_sf))

glimpse(kaz_ADM1_sf)

```

Load geographic coverage by Oblasts of the country from shapefiles in subdirectory `shape`.

```{r kaz_adm1_sf_key_count}
kaz_ADM1_sf %>%
  sf::st_drop_geometry() %>% 
    dplyr::count(ADM1_PCODE, ADM1_EN, ADM1_KK, ADM1_RU, ISO_3166_2)
```

With these Oblast keys you can easily filter Oblasts for plotting or you can list different regional (Multi-Oblast) breakdowns.

```{r plot map, fig.height = 4, fig.width = 7}

kaz_ADM1_sf <- 
  kaz_ADM1_sf %>% 
    dplyr::mutate(Regions = dplyr::case_when(
      ADM1_PCODE %in% c("KZ11", "KZ39", "KZ55", "KZ59", "KZ71") ~ "North",
      ADM1_PCODE %in% c("KZ15", "KZ23", "KZ27", "KZ47")         ~ "West",
      ADM1_PCODE %in% c("KZ35", "KZ62")                         ~ "Center",
      ADM1_PCODE %in% c("KZ10", "KZ63")                         ~ "East",
      TRUE                                                      ~ "South"      
    ) %>% 
      factor(., levels = c("North", "West", "Center", "East", "South")))

tmap::qtm(
  shp = kaz_ADM1_sf, 
  fill = "Regions", 
  text = "ADM1_EN",
  fill.palette = "Set3",
  title = "Regions of Kazakhstan",
  projection = 2502L)   # Pulkovo 1942 / Gauss-Kruger CM 69E. See <https://epsg.io/2502>.

```

If necessary, a similar operation can be performed in **Python**.

```{r Config Python}
#======================================================================
# Starting Python throu package `reticulate`
#======================================================================

library('reticulate')       # Interface to 'Python'

# options(rstudio.python.installationPath = "D:/Python/Miniconda3")
# Sys.setenv(RETICULATE_MINICONDA_PATH = "D:/Python/Miniconda3")
# reticulate::install_miniconda(path = miniconda_path(), update = TRUE, force = FALSE)
# reticulate::conda_install(envname  = 'r-reticulate', packages = c('matplotlib', 'geopandas'))
reticulate::miniconda_path()
# reticulate::use_miniconda()
# reticulate::use_condaenv('D:\\Python\\Miniconda3\\envs\\r-reticulate', required = TRUE)
reticulate::py_discover_config()

# reticulate::conda_list()  # if conda is not exist Error
```

If the required packages are not available, then they should be installed in the required ``Python`` instance. **{GeoPandas}** depends on the following packages:

* ``pandas`` - operations with dataframe.
* ``shapely`` - analysis and manipulation of geometry features.
* ``fiona`` -  read and write dataframe using multi-layered GIS formats.
* ``pyproj`` - Python interface to PROJ (cartographic projections and coordinate transformations library).

``fiona``, in turn, depends on ``attrs``, ``click``, ``cliji``, ``click_plugins``, ``munch`` and, of couse, ``GDAL`` packages.

```{bash Install libs into Python, eval=FALSE, include=FALSE}
# ```{bash}

D:
cd D:/Python/Miniconda3/envs/r-reticulate
activate r-reticulate
pip install matplotlib
pip install geopandas

pip show GDAL fiona attrs pyproj shapely geopandas matplotlib
# pip install --trusted-host pypi.python.org --trusted-host pypi.org --trusted-host files.pythonhosted.org GDAL fiona attrs pyproj shapely geopandas matplotlib

```

Further, ``matplotlib`` is an optional dependency, required for plotting, and [``rtree``](https://github.com/Toblerity/rtree) is an optional dependency, required for spatial joins. ``rtree`` requires the C library [``libspatialindex``](https://github.com/libspatialindex/libspatialindex).

Let's open the **Python** packages required for this task and read shapefiles.

```{python Load Libs & Data}
# ```{python}
# , engine.path="D:/Python/Miniconda3/envs/r-reticulate/python.exe"

#======================================================================
# Loading libraries in Python
#======================================================================

import sys
print(f'\nPython Ver. =  {sys.version}')

# importing necessary libraries
# import numpy as np
import pandas as pd
import geopandas as gpd

gpd.show_versions()

# import requests
# import uuid
# import fiona
# from osgeo import gdal
# 
# request = requests.get('https://github.com/OSGeo/gdal/blob/trunk/autotest/ogr/data/poly.zip?raw=true')
# vsiz = '/vsimem/{}.zip'.format(uuid.uuid4().hex) # gdal/ogr requires a .zip extension
# 
# gdal.FileFromMemBuffer(vsiz,bytes(request.content))
# with fiona.Collection(vsiz, vsi='zip', layer ='poly') as f:
#     gdf = gpd.GeoDataFrame.from_features(f, crs=f.crs)
#     print(gdf.head())

#======================================================================
# Loading geographic coverage in Python
#======================================================================

kaz_ADM1_gdf = gpd.read_file(r.shpfile, driver = 'ESRI Shapefile', encoding = 'utf-8')

# # open the file
# file = open('https://github.com/arodionoff/geokz/blob/main/inst/shape/kaz_admbnda_adm1_2018.zip', 'rb')
# # see https://geopandas.readthedocs.io/en/latest/docs/user_guide/io.html#reading-spatial-data
# kaz_ADM1_gdf = gpd.read_file(file)

# kaz_ADM1_gdf = gpd.read_file('/vsicurl/https://github.com/arodionoff/geokz/blob/main/inst/shape/kaz_admbnda_adm1_2018.shp?raw=true')

# Get Projection of GeoDataFrame
kaz_ADM1_gdf.crs
print(kaz_ADM1_gdf)

```

Let's open the **Python**'s ``matplotlib`` package required for visualization and show a map of the regions of Kazakhstan.

```{python Show Map}
# ```{python}
import matplotlib.pyplot as plt
import matplotlib.patches as patches

# Categories in legends produced by geopandas are sorted and this is hardcoded
# https://stackoverflow.com/questions/54370302/changing-the-order-of-entries-for-a-geopandas-choropleth-map-legend

# Set Number Code for Regions and Customer Legend
kaz_ADM1_gdf.loc[(kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ11') | 
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ39') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ55') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ59') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ71'), 'Regions'] = 0    # 'North'
kaz_ADM1_gdf.loc[(kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ15') | 
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ23') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ27') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ47'), 'Regions'] = 1    # 'West'
kaz_ADM1_gdf.loc[(kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ35') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ62'), 'Regions']   = 2  # 'Center'
kaz_ADM1_gdf.loc[(kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ10') | 
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ63'), 'Regions']   = 3  # 'East'
kaz_ADM1_gdf.loc[(kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ19') | 
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ31') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ33') | 
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ43') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ61') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ75') |
                 (kaz_ADM1_gdf['ADM1_PCODE'] == 'KZ79'), 'Regions'] = 4  # 'South'

#Re-Projection
kaz_ADM1_gdf = kaz_ADM1_gdf.to_crs(epsg = 2502)  # Pulkovo 1942 / Gauss-Kruger CM 69E. See <https://epsg.io/2502>
kaz_ADM1_gdf.crs

# Create color dictionaries for Regions- Set colors for Customer Palette in R: tmaptools::get_brewer_pal("Set3", n = 5)
palette_symb = {0: '#8DD3c7', 1: '#FFFFB3', 2: '#BEBADA', 3: '#FB8072', 4: '#80B1D3'}
regions_symb={0: 'North', 1: 'West', 2: 'Center', 3: 'East', 4: 'South'}

# Plot Geographic Coverage
fig, ax = plt.subplots(figsize=(7, 4))

kaz_ADM1_gdf.plot(
  # column = 'Regions',
  # cmap = 'Set3',
  color = kaz_ADM1_gdf['Regions'].map(palette_symb),
  edgecolor = 'black',
  linewidth = 0.15,
  categorical = True,
  legend = False,
  ax = ax
  )

kaz_ADM1_gdf.apply(lambda x: ax.annotate(text=x.ADM1_EN, xy=x.geometry.centroid.coords[0], ha='center', size=9), axis=1)

# Create Customer Legend for Category Legend
# see https://matplotlib.org/stable/gallery/text_labels_and_annotations/custom_legends.html
legend_elements = []
for x in range(len(regions_symb)):
    legend_elements.append(patches.Patch(facecolor=palette_symb[x],
                                         edgecolor='black',
                                         label=regions_symb[x]))
    
ax.legend(handles=legend_elements, bbox_to_anchor=(1.1, 1.1), title='Regions', prop={'size': 9})
ax.set(title = 'Regions of Kazakhstan')
ax.set_axis_off()

plt.show()

```
